Timid tides or tall tails? \\ Investigating the outer reaches of the Sculptor and Ursa Minor dwarf galaxies

### 2025-04-24 thesis/data.tex
tests to validate gaia filters and thesis updates

J+24 select stars initially from Gaia within a 2--4 degree circular
region centred on the dwarf satisfying:

\begin{itemize}
\tightlist
\item
  Solved parallax, proper motions, colour, and magnitudes.
\item
  High quality astrometry (\texttt{ruwe\ \textless{}=\ 1.3})
\item
  3\(\sigma\) consistency of measured parallax with dwarf distance +
  uncertainty (typically near zero; with \citet{lindegren+2018}
  zero-point correction).
\item
  Absolute RA and Dec proper motions less than
  10\(\,{\rm mas\ yr^{-1}}\) (corresponding to tangental velocities of
  \(\gtrsim 500\) km/s at distances larger than 10 kpc.)
\item
  No colour excess (\citet{lindegren+2018} equation C.2)
\item
  G \textless{} 22 and less than 5\(\sigma\) above TRGB, and between
  -0.5 and 2.5 in Bp - Rp.
\end{itemize}

Photometry is dereddened using \citet{schlegel+1988} extenction.

J+24 calculate the probability that any star belongs to either the
satellite or the MW background as \[
P_{\rm sat} = \frac{f_{\rm sat}{\cal L}_{\rm sat}}{f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}}
\]

where the satellite (sat) and background (bg) likelihoods are simply the
product of the PM, CMD, and spatial components: \[
{\cal L} = {\cal L}_{\rm space}\ {\cal L}_{\rm PM}\ {\cal L}_{\rm CMD}
\]

The satellite likelihood is constructed as

@@

The background likelihood is constructed as:

\begin{itemize}
\tightlist
\item
  CMD : Constructed as a density map using the other quality-selected
  stars outside of \(5R_h\) in the catalogue. The map is a sum of
  bivariate gaussians for each star with standard deviations based on
  the Gaia uncertainties.
\item
  PM: same as CMD except in PM space.
\item
  Spatial: a constant likelihood.
\end{itemize}

Note that each likelihood map is normalized over the respective 2D
parameter space. In order to represent the difference in frequency of
background and forground stars, \(f_{\rm sat}\) represents the field
fraction of member stars.

In J+24, a MCMC simulation is ran using the above likelihood to solve
for the following parameters

@@

We adopt a probability cut of \(P_{\rm sat} = 0.2\) as our fiducial
sample. Most stars are assigned probabilities close to 0 or 1, so the
choice of probability threshhold is not too significant. Additionally,
even for a probability cut of 0.2, the purity of the resulting sample
with RV measurements is very high (\textasciitilde90\%, J+24). (Note:
there is likely a high systematic bias in using stars with RV
measurements to measure purity. Fainter stars have poorer measurements
and distant stars are less likely to have been targeted. )

\subsection{Resulting Samples}\label{resulting-samples}

In figures fig.~\ref{fig:sculptor_selection},
fig.~\ref{fig:umi_selction}, we illustrate the resulting samples from
the algorithm. In each case, each criteria plays a roll: proper motions
are centred around the dwarf systemic motion, the CMD is well defined,
and stars only within a few \(R_h\) are included. We also plot the RV
members found in general and in \citet{sestito+2024},
\citet{sestito+2024b}.

The tangent plane, \(\xi\), \(\eta\), is the projection that

We also illustrate the approximate result of removing the spatial
component from the likelihood. We define the CMD+PM selection as stars
satisfying \[
{\cal L}_{\rm CMD,\ sat}\ {\cal L}_{\rm PM,\ sat} > {\cal L}_{\rm CMD,\ bg}\ {\cal L}_{\rm PM,\ bg}
\]

These stars appear similar to the fiducial (probable members) sample,
but instead also appear as an approximately uniform distribution across
the entire field. This illustrates the approximate background of stars
which may be confused as members. Additionally, since there is no clear
spatial structure in the CMD+PM sample, it is unlikely that there are
tidal tails detectable with Gaia. Not shown here, we also try a variety
of simpler, absolute cuts and thresholds, finding no extended structure
beyond what is detected in J+24.

This means that at least at the level of where the background density
dominates, we can exclude models which produce tidal tails brighter than
a density of
\(\Sigma_\star \approx 10^{-2}\,\text{Gaia-stars\ arcmin}^{-2} \approx 10^{-6} \, {\rm M_\odot\ kpc^{-2}}\)
(TODO assuming a distance of \ldots{} and stellar mass of \ldots).

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_selection.pdf}}
\caption[Sculptor selection criteria]{The selection criteria for Scl
members. Probable members (2-component) are orange, and all field stars
(satisfying quality criteria) are in light grey. \textbf{Top:} Tangent
plane. \textbf{Bottom left:} Colour magnitude diagram. \textbf{Bottom
right:} Proper motion.}\label{fig:sculptor_selection}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_selection.pdf}}
\caption[Ursa Minor Selection]{Similar to
fig.~\ref{fig:sculptor_selection} except for Ursa Minor. UMi features a
very extended density profile with some stars \textasciitilde{} 6\(R_h\)
including a RV member. UMi is also highly elliptical compared to other
classical dwarfs.}\label{fig:umi_selection}
\end{figure}

@@

Our primary observational constraint is the density profile of a dwarf
galaxy.

To derive density profiles, we use 0.05 dex bins in log radius (i.e.~the
bins are derived from 10\^{}(minimum(logR):0.05:maximum(logR))). The
density in each bin is then (from Poisson statistics) \[
\Sigma_b = N_{\rm memb} / A_{\rm bin} \pm \sqrt{N_{\rm memb}} / A_{\rm bin}
\] where \(N_{\rm memb}\) is the number of members in the bin and
\(A_{\rm bin}\) is the area of the bin's annulus in 2D. As discussed
below, these uncertainties underrepresent the true uncertainty on
multiple accounts. We retain Poisson errors for simplicity here.

In Figures fig.~\ref{fig:scl_observed_profiles},
fig.~\ref{fig:umi_observed_profiles}, we show the derived density
profiles for each galaxy for samples similar to in the selection plots
above. In each case, all samples are the same towards the inner regions
of the satellite, illustrating that these density profiles are dominated
by satellite stars in the centre. The sample containing all stars
reaches a plateau at the total background in the field. However,
restricting stars to being most likely satellite members by CMD + PM,
the background is much lower. This plateau likely represents the real
background of background stars which could be mistaken as members.
Finally, we have the probable members and bg-subtracted densities. BG
subtracted is based on the \texttt{all} density profile, subtracting the
mean background density for stars beyond the last point of the BG
subtracted profile.

Note that the probable members (fiducial) density profile continues to
confidently estimate the density profile below the CMD+PM likely star
background. These points are likely unreliable (see discussion below).
However, before this point, both the BG subtracted and probable members
density profiles are strikingly similar. Assumptions about the details
of the likelihood and spatial dependence have marginal influence on the
resulting density profile when the satellite is higher density than the
background.

@@

\section{Comparison of the Classical
dwarfs}\label{comparison-of-the-classical-dwarfs}

Classical dwarfs are often the brightest dwarfs in the sky.c The density
profiles of classical dwarf galaxies is thus well measured, enabling
detailed comparisons.

\begin{table*}[t]
\centering
\begin{tabular}{lll}
\toprule
galaxy & R\_h (exp inner 3Rs) & num cand\\
\midrule
Fornax & $17.8\pm0.6$ & 23,154\\
Leo I & $3.7\pm0.2$ & 1,242\\
Sculptor & $9.4\pm0.3$ & 6,888\\
Leo II & $2.4\pm0.3$ & 347\\
Carina & $8.7\pm0.4$ & 2,389\\
Sextans I & $20.2 \pm0.9$ & 1,830\\
Ursa Minor & $11.7 \pm 0.5$ & 2,122\\
Draco & $7.3\pm0.3$ & 1,781\\
\bottomrule
\end{tabular}
\end{table*}

Using the same methods above, we select members from J+24's stellar
probabilities. We use the one-component exponential density profiles
with

Figure: Density profiles for each dwarf galaxy. Here, we use the
1-component exponential stellar probabilities from J+24. Dwarf galaxies
are scaled by our derived \(R_h\) values.

\subsection{Density Profile Reliability and
Uncertainties}\label{density-profile-reliability-and-uncertainties}

\begin{itemize}
\tightlist
\item
  How well do we know the density profiles?
\item
  What uncertainties affect derived density profiles?
\item
  Can we determine if Gaia, structural, or algorithmic systematics
  introduce important errors in derived density profiles?
\item
  Using J+24 data, we validate

  \begin{itemize}
  \tightlist
  \item
    Check that PSAT, magnitude, no-space do not affect density profile
    shape too significantly
  \end{itemize}
\item
  Our ``high quality'' members all have \textgreater{} 50 member stars
  and do not depend too highly on the spatial component, mostly
  corresponding to the classical dwarfs
\end{itemize}

J+24's algorithm takes spatial position into account, assuming either a
one or two component exponential density profile. When deriving a
density profile, this assumption may influence the derived density
profile, especially when the galaxy density is fainter than the
background of similar appearing stars. To remedy this and estimate where
the background begins to take over, we also explore a cut based on the
likelihood ratio of only the CMD and PM components. This is in essence
assuming that the spatial position of a star contains no information on
it's membership probability (a uniform distribution like the background)

@@

The J+24 method was designed to determine high probability members for
spectroscopic followup in particular. Note that we instead care about
retrieving a reliable density profile.

In particular, in fig.~\ref{fig:umi_observed_profiles}, notice that the
PSAT method produces artifically small errorbars even when the density
is \textgreater1dex below the local background. These stars are likely
selecting stars from the statistical MW background consistent with UMi
PM / CMD, recovering the assumed density profile. As a result, the
reliability of faint features in these density profiles is questionable
and a more robust analysis, removing this particular density assumtion,
would be required to more appropriately represent the knowledge of the
density profile as the background begins to dominate.

J+24 do not account for structural uncertainties in dwarfs. This is a
not insignificant source of uncertainty in the derived density profile

We assume constant ellipticity and position angle. Dwarf galaxies, in
reality, are not necessarily smooth and constant.

In summary, we have used J+24 data to derive the density profiles for
Fornax, Sculptor, and Ursa Minor. In each case, the density profile is
robust against different selection criteria. Both Sculptor (Ursa Minor)
show strong (weak) evidence for deviations from an exponential profile.
We will explore a tidal explanation for these features in this work.

@@

For both Sculptor and Ursa Minor, we construct samples of radial
velocity measurements and cross match each sample to produce estimates
of the total velocity dispersion and line of sight velocity for each
galaxy.

\subsection{Methodology}\label{methodology}

@@

We combine the mean RV measurement from each study using the
inverse-variance weighted mean and standard uncertainty. \[
\bar v = \frac{1}{\sum w_i}\sum_i w_i\ v_{i} \\
\delta v = \sqrt{\frac{1}{\sum_i w_{i}^2}}
\] where \(w_i = 1/s_i^2\), and we estimate the inter-study standard
deviation \[
s^2 = \frac{1}{\sum w_i} \sum_i w_i (v_{i} - \bar v)^2
\] We remove stars with significant velocity dispersions as measured
between or within a study: \[
s < 5\,\delta v\,\sqrt{n}
\] where \(s, \delta v, n\) are the standard deviation, standard, error,
and number of measurements for both a study (with multiple epochs) and
inter-study comparisons.

The combined RV likelihood is then \[
{\cal L} = {\cal L}_{\rm space} {\cal L}_{\rm CMD} {\cal L}_{\rm PM} {\cal L}_{\rm RV}
\] where \[
{\cal L}_{\rm RV, sat} = N(\mu_{v}, \sigma_{v}^2 + (\delta v_i)^2) \\
{\cal L}_{\rm RV, bg} = N(0, \sigma_{\rm halo}^2)
\] where \(\mu_v\) and \(\sigma_v\) are the systemic velocity and
dispersion, and \(\delta v_i\) is the individual measurement
uncertainty. Typically, the velocity dispersion will dominate the
uncertainty budget here. We assume a halo/background velocity dispersion
of a constant \$\sigma\_\{\rm halo\} = 100 \$ km/s
\citep[e.g.][]{brown+2010}.

@@

Finally, we need to correct the coordinate frames for the solar motion
and on-sky size of the galaxy. The first step is to subtract out the
solar motion from each radial velocity, corresponding to a typical
gradient of \textasciitilde3 km/s across the field. The next step is to
account for the slight differences in the direction of each radial
velocity. Define the \(\hat z\) direction to point parallel to the
direction from the sun to the centre of the dwarf galaxy. Then if
\(\phi\) is the angular distance between the centre of the galaxy and
the individual star, the corrected radial velocity is then \[
v_z = v_{\rm los, gsr}\cos\phi  - v_{\alpha}\cos\theta \sin\phi - v_\delta \sin\theta\sin\phi
\] where
\(v_{\rm tan, R} = d(\mu_{\alpha*}\cos\theta + \mu_\delta \sin \theta)\)
is the radial component of the proper motion with respect to the centre
of the galaxy.This correction is of the order \(v_{\rm tan}\theta\) so
induces a gradient of about \(1 km/s/degree\) for sculptor
\citep[see][]{WMO2008}. The uncertainty is then the velocity uncertainty
plus the distance uncertainties times the PM uncertainty from above. We
then use the \(v_z\) values for the following modelling, however
repeating with plain, solar-frame velocities does not substantially
affect the results too much .

@@

@@

\begin{table*}[t]
\centering
\caption{Summary of velocity measurements and derived properties.}
\label{tbl:Summary-of-velocity-measurements-and-derived-properties}
\begin{tabular}{llllllll}
\toprule
 & Study & Instrument & Memb & Rep. & N$\sigma > s$ & $\bar v$ & $\sigma_v$\\
\midrule
Scl &  &  &  &  &  &  & \\
 & tolstoy+23 & VLT/FLAMES & 200 (1604) & 500 &  &  & 9.6\\
 & sestito+23a & GMOS & 2000 &  &  &  & -\\
 & walker+09 & MMFS & 2000 &  &  &  & 9.5\\
 & APOGEE & APOGEE & 200 & 200 &  & 8 & 9\\
UMi &  &  &  &  &  &  & \\
 & sestito+23b & GRACES & 6 &  &  &  & \\
 & pace+20 & Keck/DEIMOS &  &  &  & -245 & 8.6\\
 & spencer+18 & MMT/Hectoshell &  &  &  & -247 & \\
 & APOGEE & APOGEE & 100 & 100 &  &  & \\
\bottomrule
\end{tabular}
\end{table*}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_vlos_xi_eta.pdf}}
\caption[Scl velocity sample]{A plot of the corrected los velocities for
Scl binned in tangent plane coordinates. We detect a slight rotational
gradient towards the bottom right. \textbf{TODO} add a histogram of
velocities here too with fits, maybe with respect to distance.}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_vel_gradient.pdf}}
\caption[Scl velocity gradient]{A velocity gradient in Sculptor! The
arrow marks the gradient induced by Scl's proper motion on the sky.
\textbf{TODO}: plot of velocity of Scl in direction of gradient.}
\end{figure}

@@

@@

For UMi, we collect radial velocities from, APOGEE,
\citet{sestito+2024b}, \citet{pace+2020}, and \citet{spencer+2018}.

We shifted the velocities of \citet{spencer+2018} (\(-0.9\) km/s) and
\citet{pace+2020} (\(+1.0\) km/s) to reach the same scale based on a
crossmatch of about 200 common stars. Since the mean difference in
velocities in this crossmatch is of order 1.9 km/s, we adopt this as the
systematic LOS velocity error.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_vlos_xi_eta.pdf}}
\caption[UMi velocity sample]{A plot of the corrected los velocities for
UMi binned in tangent plane coordinates. There is no clear rotation or
velocity gradient here. Interestingly, many velocity members are as far
as 100 arcmin away.}
\end{figure}

@@

\begin{itemize}
\tightlist
\item
  Inter-study systematics and biases. While basic crossmatches and a
  simple velocity shift, combining data from multiple instruments is
  challenging.
\item
  Inappropriate uncertainty reporting. Inspection of the variances
  compared to the standard deviations within a study seems to imply that
  errors are accurately reported. APOGEE notes that their RV
  uncertainties are known to be underestimates.
\item
  Binarity. While not too large of a change for classical dwarfs, this
  could inflate velocity dispersions of about 9 km/s by about 1 km/s
  \citet{spencer+2017}. Thus, our measurement is likely slightly
  inflated given the high binarity fractions measured in classical
  dwarfs \citep[\citet{spencer+2018}]{arroyo-polonio+2023}.
\item
  Selection effects. RV studies each have their own selection effects. I
  do not know how to correct for this.
\end{itemize}

>>>>

J+24 select stars initially from Gaia within a 2--4 degree circular
region centred on the dwarf. J+24 only consider stars with:

\begin{itemize}
\tightlist
\item
  Solved parallax, proper motions, colour, and magnitudes.
\item
  High quality astrometry (\texttt{ruwe\ \textless{}=\ 1.3}).
\item
  3\(\sigma\) consistency of measured parallax with dwarf distance +
  uncertainty (typically near zero; with \citet{lindegren+2018}
  zero-point correction).
\item
  Absolute RA and Dec proper motions less than
  10\(\,{\rm mas\ yr^{-1}}\) (corresponding to tangental velocities of
  \(\gtrsim 500\) km/s at distances larger than 10 kpc.).
\item
  No colour excess (\citet{lindegren+2018} equation C.2).
\item
  22 \textgreater{} G \textgreater{} 5\(\sigma\) brighter than TRGB (in
  practice astrometry ruwe =\textgreater{} 21 \textgreater{} G).
\item
  Between -0.5 and 2.5 in Bp - Rp.
\end{itemize}

Photometry is dereddened using \citet{schlegel+1988} extinction maps.

J+24 calculate the probability that any star belongs to either the
satellite or the MW background with \[
P_{\rm sat} = \frac{f_{\rm sat}{\cal L}_{\rm sat}}{f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}}
\]

where \(f_{\rm sat}\) is the fraction of candidate members in the field,
and \({\cal L}_{\rm sat}\) and \({\cal L}\)\_\{\rm bg\}\$ are the
satellite (sat) and background (bg) likelihoods respectively. Each
likelihood is the product of a spatial, proper motion, and CMD term: \[
{\cal L} = {\cal L}_{\rm space}\ {\cal L}_{\rm PM}\ {\cal L}_{\rm CMD}.
\]

The satellite likelihoods are specified as

@@

The background likelihoods are constructed as:

\begin{itemize}
\tightlist
\item
  CMD : Constructed as a density map using the other quality-selected
  stars outside of \(5R_h\) in the catalogue. The map is a sum of
  bivariate gaussians for each star with standard deviations based on
  the Gaia uncertainties.
\item
  PM: same as CMD except in PM space (\(\mu_{\alpha*}\) and
  \(\mu_\delta\) with respecive uncertainties and covariance.)
\item
  Spatial: a constant likelihood.
\end{itemize}

Each likelihood map is normalized over the respective 2D parameter
space. \(f_{\rm sat}\) thus is the only term controlling the total
abundance of satellite stars relative to the background.

In J+24, a MCMC simulation is ran using the above total likelihood to
solve for the following parameters:

@@

We adopt a probability cut of \(P_{\rm sat} = 0.2\) as our fiducial
sample. Most stars are assigned probabilities close to 0 or 1, so the
choice of probability threshhold is not too significant. Additionally,
even for a probability cut of 0.2, the purity of the resulting sample
with RV measurements is very high (\textasciitilde90\%, J+24). However,
there is likely a systematic bias in using stars with RV measurements to
measure purity. Fainter stars have poorer astrometry and are less likely
to have been targeted. We find no difference in the resulting density
distributions when restricting stars to be brighter than a specific
magnitude.

\subsection{Selected samples}\label{selected-samples}

In figures fig.~\ref{fig:sculptor_selection},
fig.~\ref{fig:umi_selction}, we illustrate the resulting samples from
the algorithm in the tangent plane (\(\xi\), \(\eta\), \emph{does this
need defined?}), CMD, and proper motion space. For both galaxies, each
criteria plays a commensurate role in sifting out members. Proper
motions are clustered around the dwarf systemic motion, the CMD is well
defined including the horizontal branch, and stars only within a few
\(R_h\) are included. We also mark stars with consistent radial
velocities (RVs) (see below), which trace each feature albeit more
cautiously.

Figure REF also illustrats the distribution of stars selected without a
spatial criterion. We define the CMD+PM selection as stars satisfying \[
{\cal L}_{\rm CMD,\ sat}\ {\cal L}_{\rm PM,\ sat} > {\cal L}_{\rm CMD,\ bg}\ {\cal L}_{\rm PM,\ bg}
\] These stars are distributed similar to the fiducial (probable
members) sample but with a fainter uniform distribution across the
entire field. This illustrates the approximate background of stars which
may be confused as members. Additionally, since there is no clear
spatial structure in the CMD+PM sample, it is unlikely that there are
additional faint, tidal features detectable with Gaia. Not shown here,
we also try a variety of simpler, absolute cuts and thresholds, finding
no extended structure beyond what is detected in J+24. (\emph{Is it
worth using this to calculate an OOM upper limit on the density of tidal
tails?}).

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_selection.pdf}}
\caption[Sculptor selection criteria]{The selection criteria for Scl
members. Probable members (2-component) are orange, and all field stars
(satisfying quality criteria) are in light grey. \textbf{Top:} Tangent
plane. We outline in star symbols the two stars from
\citet{sestito+2023a}. \textbf{Bottom left:} Colour magnitude diagram.
\textbf{Bottom right:} Proper motion.}\label{fig:sculptor_selection}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_selection.pdf}}
\caption[Ursa Minor Selection]{Similar to
fig.~\ref{fig:sculptor_selection} except for Ursa Minor. We outline RV
members outside of \(3R_h\) in black stars (from \citet{sestito+2023b}
and \citet{pace+2020} and
\citet{spencer+2018}).}\label{fig:umi_selection}
\end{figure}

@@

Density profiles are an essential observational constraint for our later
simulations. To derive density profiles, we use 0.05 dex bins in log
radius (i.e.~the bins are derived from
10\^{}(minimum(logR):0.05:maximum(logR))). We remove bins at smaller
(larger) radii than the first bin to contain no stars, working outwards.
We use poisson uncertainties
(\(\delta \Sigma_i / \Sigma_i = 1/\sqrt{N_i}\) for a derived density
\(\Sigma_i\) in a bin with \(N_i\) stars). As discussed below, these
uncertainties are straightforward but are likely under-represented.

fig.~\ref{fig:scl_observed_profiles},
fig.~\ref{fig:umi_observed_profiles} show the derived density profiles
for Scl and UMi. We calculate density profiles for different selections
of stars from above: all quality stars, CMD+PM only, and the probable
member samples. In each case, all samples are the same towards the inner
regions of the satellite. Classical dwarfs dominate the stellar density
in their cores. However, the density profile \emph{all stars} plateaus
at the total background in the field at radii of 30-60 arcminutes. By
restricting stars to being most likely satellite members by CMD + PM,
the background is reduced by 1-2 dex. The CMD+PM background plateau
likely represents the density of background stars which could be
mistaken as members. Finally, the background (BG)-subtracted profile
results from subtracting the apparent background in the \emph{all}
profile.

Note that the probable members (fiducial) density profile continues to
confidently estimate the density profile below the CMD+PM background.
These points are likely unreliable (see discussion below). However,
before this point, both the BG subtracted and probable members density
profiles are strikingly similar. Assumptions about the details of the
likelihood and spatial dependence have marginal influence on the
resulting density profile when the satellite is higher density than the
background. Thus the detection of excesses of stars in Sculptor (past 20
arcmin) and UMi (past 15-20 arcmin) are robust.

One interesting note is for UMi, there is a small ultrafaint star
cluster at a position of (-100, -15) arcmin \citet{munoz+2012}. However
this cluster does not have a bright RGB so has few stars brighter than g
mag of 22. While munoz 1 may contribute about 4 stars to the density
profile of UMi, this would only apply to the bin at XXX because of the
small half light radius. We do not subtract this out.

@@

@@

The J+24 method was designed in particular to detect the presence of a
density excess and find individual stars at large radii to be followed
up. We are more interested in accurately quantifying the density profile
and size of any perturbations. One potential problem with using J+24's
candidate members is that the algorithm assumes the density is either
described by a single or double exponential. If this model does not
accurately match the actual density profile of the dwarf galaxy, we
would like to understand how strongly influenced the density profile is
by this assumption.

In particular, in fig.~\ref{fig:umi_observed_profiles}, notice that the
PSAT method produces artifically small errorbars even when the density
is \textgreater1dex below the local background. These stars are likely
selecting stars from the statistical MW background consistent with UMi
PM / CMD, recovering the assumed density profile. As a result, the
reliability of these density profiles below the CMD+PM background may be
questionable. A more robust analysis, removing this particular density
assumption, would be required to more appropriately represent the
knowledge of the density profile as the background begins to dominate.

J+24 do not account for structural uncertainties in dwarfs for the two
component case. We assume constant ellipticity and position angle. Dwarf
galaxies, in reality, are not necessarily smooth and constant. They do
test an alternative method using circular radii for the extended density
component, and we find these density profiles are very similar. We also
assume a constant ellipticity and position angle here.

@@

For both Sculptor and Ursa Minor, we construct literature samples of
radial velocity measurements. We combine these samples with J+24's
members to produce RV consistent stars and to compute velocity
dispersion, systematic velocities, and test for the appearance of
velocity gradients.

@@

We combine the mean RV measurement from each study using the
inverse-variance weighted mean and standard uncertainty. \[
\bar v = \frac{1}{\sum w_i}\sum_i w_i\ v_{i} 
\] \[
\delta v = \sqrt{\frac{1}{\sum_i w_{i}^2}}
\]

where \(w_i = 1/s_i^2\), and we estimate the inter-study standard
deviation \[
s^2 = \frac{1}{\sum w_i} \sum_i w_i (v_{i} - \bar v)^2
\] We remove stars with significant velocity dispersions as measured
between or within a study: \[
P\left(\chi2(N-1) < \frac{s^2}{\delta v^2}\right) > 0.001
\] where \(s, \delta v, N\) are the weighted standard deviation,
weighted standard error, and number of observations. We apply this cut
to both within a study and between studies, which typically removes
stars with reduced \(\chi^2 \gtrsim 2\).

The combined RV likelihood is then \[
{\cal L} = {\cal L}_{\rm space} {\cal L}_{\rm CMD} {\cal L}_{\rm PM} {\cal L}_{\rm RV}
\] where \[
{\cal L}_{\rm RV, sat} = N(\mu_{v}, \sigma_{v}^2 + (\delta v_i)^2)
\] \[
{\cal L}_{\rm RV, bg} = N(0, \sigma_{\rm halo}^2)
\]

where \(\mu_v\) and \(\sigma_v\) are the systemic velocity and
dispersion of the satellite, and \(\delta v_i\) is the individual
measurement uncertainty. Typically, the velocity dispersion will
dominate the uncertainty budget here. We assume a halo/background
velocity dispersion of a constant \(\sigma_{\rm halo} = 100\) km/s
\citep[e.g.][]{brown+2010}.

@@

Finally, we need to correct the coordinate frames for the solar motion
and on-sky size of the galaxy. The first step is to subtract out the
solar motion from each radial velocity, corresponding to a typical
gradient of \textasciitilde3 km/s across the field. The next step is to
account for the slight differences in the direction of each radial
velocity. Define the \(\hat z\) direction to point parallel to the
direction from the sun to the centre of the dwarf galaxy. Then if
\(\phi\) is the angular distance between the centre of the galaxy and
the individual star, the corrected radial velocity is then \[
v_z = v_{\rm los, gsr}\cos\phi  - v_{\alpha}\cos\theta \sin\phi - v_\delta \sin\theta\sin\phi
\] where
\(v_{\rm tan, R} = d(\mu_{\alpha*}\cos\theta + \mu_\delta \sin \theta)\)
is the radial component of the proper motion with respect to the centre
of the galaxy.This correction is of the order \(v_{\rm tan}\theta\) so
induces a gradient of about \(1 km/s/degree\) for sculptor
\citep[see][]{WMO2008}. The uncertainty is then the velocity uncertainty
plus the distance uncertainties times the PM uncertainty from above. We
then use the \(v_z\) values for the following modelling, however
repeating with uncorrected, heliocentric velocities does not
significantly affect the results .

@@

\subsection{}\label{section}

\begin{table*}[t]
\centering
\caption{Summary of velocity measurements and derived properties. sestito+2023a number of members depends on spatial model used.}
\label{tbl:Summary-of-velocity-measurements-and-derived-properties-sestito+2023a-number-of-members-depends-on-spatial-model-used}
\begin{tabular}{llllllll}
\toprule
 & Study & Instrument & Nspec & Nstar & Ngood & Nmemb & $\delta v_{\rm med}$\\
\midrule
Scl & combined &  & 8945 & 2280 & 2034 & 1920 & 0.9\\
 & tolstoy+23 & FLAMES & 3311 & 1701 & 1522 & 1481 & 0.65\\
 & sestito+23a & GMOS & 2 & 2 & 2 & 2 & 13\\
 & walker+09 & MMFS & 1818 & 1522 & 1417 & 1330 & 1.8\\
 & APOGEE & APOGEE & 5082 & 253 & 102 & 98 & 0.6\\
UMi & combined &  & 4714 & 1225 & 1148 & 831 & 2.3\\
 & sestito+23b & GRACES & 5 & 5 & 5 & 5 & 1.8\\
 & pace+20 & DEIMOS & 1716 & 1538 & 829 & 682 & 2.5\\
 & spencer+18 & Hectoshell & 1407 & 970 & 596 & 406 & 0.9\\
 & APOGEE & APOGEE & 9500 & 279 & 37 & 32 & 0.9\\
\bottomrule
\end{tabular}
\end{table*}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_rv_2dhist.pdf}}
\caption[Scl velocity sample]{A plot of the corrected los velocities for
Scl binned in tangent plane coordinates. We detect a slight rotational
gradient towards the bottom right.}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_vel_gradient_binned.pdf}}
\caption[Scl velocity gradient]{A velocity gradient in Sculptor! The
arrow marks the gradient induced by Scl's proper motion on the sky.
\textbf{TODO}: plot of velocity of Scl in direction of gradient.}
\end{figure}

@@

Bayes factor found using relative density via KDE with Silverman's
bandwith to prior (XXX theorem).

@@

\begin{table*}[t]
\centering
\caption{MCMC fits for different RV datasets for Scl amoung 3 different models.}
\label{tbl:MCMC-fits-for-different-RV-datasets-for-Scl-amoung-3-different-models}
\begin{tabular}{lllllll}
\toprule
study & mean & sigma & $\partial \log\sigma / \partial \log R$ & $\partial v_z / \partial x$ (km/s/deg) & $\theta_{\rm grad} / ^{\circ}$ & $\log bf$\\
\midrule
all &  &  &  &  &  & \\
 & $111.17\pm0.22$ & $9.68\pm0.17$ & - & - & - & 0\\
 & $111.18 \pm 0.23$ & $9.64\pm0.16$ & - & $4.8\pm1.3$ & $-147_{-12}^{+15}$ & -3.7*\\
 & $111.15\pm0.22$ & $9.70\pm0.16$ & $0.055\pm0.021$ & - & - & -0.8\\
tolstoy+23 &  &  &  &  &  & \\
 & $111.2 \pm 0.3$ & $9.77 \pm 0.18$ & - & - & - & 0\\
 & $111.2\pm0.3$ & $9.73\pm0.18$ & – & $4.8\pm1.4$ & $-154_{-12}^{+16}$ & -2.5\\
 & $111.2 \pm 0.3$ & $9.71\pm0.18$ & $0.081 \pm 0.023$ & – & – & -3.3\\
walker+09 &  &  &  &  &  & \\
 & $111.0\pm0.3$ & $9.57\pm0.21$ & – & – & – & 0\\
 & $111.1\pm0.3$ & $9.54\pm0.21$ & - & $5.3_{-1.6}^{+1.8}$ & $-134_{-16}^{+22}$ & -2.0\\
 & $111.0\pm0.3$ & $9.61\pm0.21$ & $0.03\pm0.03$ & – & – & +1.6\\
apogee &  &  &  &  &  & \\
 & $109.9\pm0.8$ & $8.3\pm0.6$ & – & – & – & –\\
 & $109.9\pm0.8$ & $8.3\pm0.6$ & – & $6\pm3$ & $-151_{-36}^{+44}$ & +0.3\\
 & $109.9\pm0.8$ & $8.3\pm0.7$ & $0.05\pm0.08$ & – & – & +1.1\\
\bottomrule
\end{tabular}
\end{table*}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_rv_2dhist.pdf}}
\caption[UMi velocity sample]{UMi velocities}
\end{figure}

Figure: A plot of the corrected los velocities for UMi binned in tangent
plane coordinates. There is no clear rotation or velocity gradient here.
Interestingly, many velocity members are as far as 100 arcmin away.

For UMi, we collect radial velocities from, APOGEE,
\citet{sestito+2023b}, \citet{pace+2020}, and \citet{spencer+2018}.

We shifted the velocities of \citet{spencer+2018} (\(-1.1\) km/s) and
\citet{pace+2020} (\(+1.1\) km/s) to reach the same scale. We found 183
crossmatched common stars (passing 3\(\sigma\) RV cut, velocity
dispersion cut, and PSAT J+24 \textgreater{} 0.2 w/o velocities). Since
the median difference in velocities in this crossmatch is about 2.2
km/s, we adopt 1 km/s as the approximate systematic error here.

@@

\begin{table*}[t]
\centering
\caption{MCMC fits for UMi velocity dispersion.}
\label{tbl:MCMC-fits-for-UMi-velocity-dispersion}
\begin{tabular}{lllll}
\toprule
study & mean & sigma & $\log bf_{\rm sigma}$ & $\log bf_{\rm grad}$\\
\midrule
all & $-245.9\pm0.3$ & $8.76\pm0.24$ & +1.5 & +2.2\\
pace & $-244.5\pm0.4$ & $9.1\pm0.3$ & +0.2 & +1.1\\
spencer & $-246.9\pm0.4$ & $8.8\pm0.3$ & +1.8 & -0.3\\
apogee & $-248.2\pm1.6$ & $9.0_{-1.1}^{+1.3}$ & +0.8 & +0.8\\
\bottomrule
\end{tabular}
\end{table*}

@@

\begin{itemize}
\tightlist
\item
  Inter-study systematics and biases. While basic crossmatches and a
  simple velocity shift, combining data from multiple instruments is
  challenging.
\item
  Inappropriate uncertainty reporting. Inspection of the variances
  compared to the standard deviations within a study seems to imply that
  errors are accurately reported. APOGEE notes that their RV
  uncertainties are known to be underestimates but are a small
  proportion of our sample.
\item
  Binarity. While not too large of a change for classical dwarfs, this
  could inflate velocity dispersions of about 9 km/s by about 1 km/s
  \citet{spencer+2017}. Thus, our measurement is likely slightly
  inflated given the high binarity fractions measured in classical
  dwarfs \citep[\citet{spencer+2018}]{arroyo-polonio+2023}.
\item
  Selection effects. RV studies each have their own selection effects. I
  do not know how to correct for this.
\end{itemize}
